---
title: "The Statistical Dangers of Between-Species Comparisons"
author: "Jake L. Weissman"
date: "July 4, 2020"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


**Disclaimer:** This is intended to be a short introduction to the idea of phylogenetically
structured data. My main goal is to make sure you know that you have to be careful
when doing statistics on species data. If you want an introduction to
phylogenetic comparative methods (a rich subfield of evolutionary biology) I recommend
starting with [Luke Harmon's free online textbook](https://lukejharmon.github.io/pcm/) or
any of [many other texts on the topic](https://link.springer.com/book/10.1007%2F978-3-662-43550-2) (or [workshop notes online](http://lukejharmon.github.io/ilhabela/) or [here](http://www.phytools.org/Cordoba2017/)).

## Comparative Data

This lesson is focused on how to deal with "phylogenetically structured" data. For
example, imagine that we have some trait data for a set of species of bacteria, perhaps
about mean cell size and minimum generation time. It is often supposed that larger cells
are associated with rapid maximal growth rates. It seems that it would be a simple
matter to test this association - one could simply calulate the correlation between
the two variables. In fact, if you proceed without caution you run the risk of obtaining
a significant, but spurious statistical result. 

Some other examples of research questions involving phylogenetically structured data:  
  - trying to associate pathogenicity with other traits (e.g., genome size)  
  - trying to associate the traits of different species with different habitat types (e.g., do beige things live in the desert and dark green things in the forst)

The problem is that most (simple) statistical tests (parametric and non-parametric) assume
independent datapoints, and your data is anything but. Consider that the species in
your dataset share an evolutionary history, leading their traits to share a hierarchical dependency
structure. Any time you are making comparisons between sets of species, you need to 
account for phylogenetic structure. [Felsenstein (1985)](https://www.jstor.org/stable/2461605) was the
[first to really articulate the severity of this problem and provide an appropriate statistical treatment](https://doi.org/10.1086/703055) (the original linked  "phylogenetic independent constrasts"
paper is very clearly written and I highly recommend it).

I'm going to take you through this issue by generating some example phylogenetically-structured datasets in R
and showing you how statistical tests can fail on these data.

I want to emphasize that I am not just being nitpicky - it is very easy to get spurious results with
comparative data. There is a great paragraph near the end of the original Felsenstein paper:

*"Some reviewers of this paper felt that the message was  "rather nihilistic," and*
*suggested that it would be much improved if I could present a simple and robust*
*method that obviated the need to have an accurate knowledge of the phylogeny. I*
*entirely sympathize, but do not have a method that solves the problem. The best*
*we can do is perhaps to use pairs of close relatives as suggested above, although*
*this discards at least half of the data. Comparative biologists may understandably*
*feel frustrated upon being told that they need to know the phylogenies of their*
*groups in great detail, when this is not something they had much interest in*
*knowing. Nevertheless, efforts to cope with the effects of the phylogeny will have*
*to be made. Phylogenies are fundamental to comparative biology; there is no * 
*doing it without taking them into account." -  [Felsenstein (1985)](https://www.jstor.org/stable/2461605)*

Before we start you will need to make sure you have several packages installed. The `ape`
package is the general R package for dealing with phylogenies, and the `phytools`
package has a ton of extra tools for manipulating and analyzing phylogenetic data.
We will use a function in the `caper` package to check for phylogenetic signal and
a function in the `picante` package to run phylogenetic generalized least squares (PGLS).

```{r packages, results='hide', message=FALSE, warning=FALSE}
if (!require("ggplot2")){
  install.packages("ggplot2")
  require("ggplot2")
}
if (!require("ape")){
  install.packages("ape")
  require("ape")
}
if (!require("phytools")){
  install.packages("phytools")
  require("phytools")
}
if (!require("picante")){
  install.packages("picante")
  require("picante")
}
if (!require("caper")){
  install.packages("caper")
  require("caper")
}
```

## Spurious Results With Phylogenetic Structure

Let's start be generating some random data to work with. I am going to simulate
the evolution of two traits on a phylogeny, and we are going to see how often
those traits end up with correlated values **even though they will have no real
association** (other than evolving on the same tree, that is).

First let's make some phylogenies:

```{r phylogenies}
# Generate 9 random phylogenies
n_trees <- 9
n_tips <- 1000
trees <- replicate(n_trees, rcoal(n_tips), simplify = FALSE)
par(mfrow=c(3,3))
for(i in 1:n_trees){
  plot.phylo(as.phylo(trees[[i]]), show.tip.label = FALSE)
}
par(mfrow=c(1,1))
```

Now I am going to simulate the evolution of two traits on each of these phylogenies. We
don't need to get into it now, but I am using a simple [Brownian Motion](https://lukejharmon.github.io/pcm/chapter3_bmintro/) model of trait
evolution that is pretty standard in the field. It results in normally-distributed
changes in trait values along branches on the tree. Remember, I am going to let these
traits evolve **independently** on the tree.

```{r traits}
# Randomly evolve two traits on these trees
trait_df <- data.frame(Trait1 = numeric(0),
                       Trait2 = numeric(0),
                       Tree = numeric(0))
for(i in 1:n_trees){
  trait_df <- rbind(trait_df, 
                    data.frame(Trait1 = fastBM(trees[[i]]),
                               Trait2 = fastBM(trees[[i]]),
                               Tree = i))
}
ggplot(trait_df,aes(x = Trait1, y = Trait2)) + 
  geom_point() + 
  stat_smooth(method="lm") + 
  facet_wrap(vars(Tree), scales = "free")
```

Wow, based on the confidence intervals those regression lines I'd say that at least
a few of these simulated datasets have some pretty significant statistical relationships between
our two traits (you can check this yourself if you really want to). But these traits don't have any real relationship! 

What if, instead, we let our traits evolve on a "star" shaped tree, where no hierarchical
structure exists?

```{r star}
n_tips <-  10 # reduce number of tips just for plotting
star <- starTree(1:n_tips, branch.lengths=rep(1,n_tips))
plot.phylo(star, type = "fan" , show.tip.label = FALSE)
```


We see a very different pattern in our traits:

```{r startraits}
# What if we simulated trait evolution on a star phylogeny instead?
n_tips <-  1000
star <- starTree(1:n_tips, branch.lengths=rep(1,n_tips))
star_df <- data.frame(Trait1 = numeric(0),
                       Trait2 = numeric(0),
                       Tree = numeric(0))
for(i in 1:n_trees){
  star_df <- rbind(star_df, 
                    data.frame(Trait1 = fastBM(star),
                               Trait2 = fastBM(star),
                               Tree = i))
}
ggplot(star_df,aes(x = Trait1, y = Trait2)) + 
  geom_point() + 
  stat_smooth(method="lm") + 
  facet_wrap(vars(Tree))
```

This time, because there is no phylogenetic structure, we don't end up with a
spurious result.

But why does phylogenetic structure lead us to a spurious result? There's a few
ways to conceptualize this but I've always found the easiest to be the idea of "pseudoreplication".
Pseudoreplication means you actually have fewer data points than you think you do. 

It helps to imagine actual replication, where maybe you've accidentally duplicated your dataset,
say 100 times.

```{r replication}
x <- rnorm(10)
y <- rnorm(10)
plot(x,y)
cor.test(x,y)

plot(rep(x,100),rep(y,100))
cor.test(rep(x,100),rep(y,100))
```

By inflating the number of datapoints you have, you can force a significant result (see [Felsenstein (1985)](https://www.jstor.org/stable/2461605)
for a discussion of this in terms of degrees of freedom - same idea).

When you draw from a phylogeny, you are essentially replicating pieces of evolutionary
history (you are double counting the times where two species were actually the same
ancestral species) - hence "pseudoreplication". Changes in trait values that happened 
early on in evolutionary history get propagated to many ofspring, artificially inflating the 
importance of these changes.

It is easiest to see this when looking at a balanced tree with two very diverged groups.

```{r balanced}
n_tips <- 2^10
balanced <- stree(n_tips, type = "balanced", tip.label = NULL)
balanced$edge.length <- rep(1,nrow(balanced$edge))
balanced$edge.length[which(balanced$edge[,1]==1025)] <- 100
plot(balanced, show.tip.label = FALSE)
```

When you simulate traits on this tree you get two distinct clusters

```{r balanced traits}
balanced_df <- data.frame(Trait1 = numeric(0),
                       Trait2 = numeric(0),
                       Tree = numeric(0))
for(i in 1:n_trees){
  balanced_df <- rbind(balanced_df, 
                    data.frame(Trait1 = fastBM(balanced),
                               Trait2 = fastBM(balanced),
                               Tree = i))
}
ggplot(balanced_df,aes(x = Trait1, y = Trait2)) + 
  geom_point() + 
  stat_smooth(method="lm") + 
  facet_wrap(vars(Tree))
```

Now, we really only have two datapoints here (our two major phylogenetic groups),
but we have created the illusion of having over 1000, and this leads to our spurious results.
Note that if you looked within either of our two clusters we would be unlikely to find any real
trend. This idea is similar to that of [Simpson's Paradox](https://en.wikipedia.org/wiki/Simpson%27s_paradox), which you should also be wary of when aggregating data that might have underlying (non-phylogenetic) structure.

## Testing for Phylogenetic Signal

Often researchers will want to know if their trait has significant phylogenetic
signal (for very fast-evolving traits evolutionary history may get "washed-out",
making correcting for it unnecessary - though some will argue you should use phylogenetically
corrected methods no-matter what).

There are a few ways to do this, but we will use two common approaches ([the non-parametric 
Bloomberg's K, which basically shuffles your tips, and the more traditional Pagel's 
lambda](http://blog.phytools.org/2012/03/phylogenetic-signal-with-k-and.html)) implemented
in the `caper` package. 

Let's test for phylogenetic signal in the traits simulated on one of our first
phylogenies:

```{r phylosig}
#Let's just look at the first tree
traits_tree1 <- subset(trait_df, trait_df$Tree == 1)
tree1 <- trees[[1]]

phylosig(tree1, traits_tree1$Trait1, method = "K", test = TRUE)
phylosig(tree1, traits_tree1$Trait1, method = "lambda", test = TRUE)
phylosig(tree1, traits_tree1$Trait2, method = "K", test = TRUE)
phylosig(tree1, traits_tree1$Trait2, method = "lambda", test = TRUE)
```

Looks like there's lots of phylogenetic signal in both of our simulated traits, as expected.

## Correcting for Phylogenetic Signal

There are lots of approaches and opinions on how to correct for phylogeny when, say,
performing a regression. The oldest approach is Felsenstein's independent
contrasts - where he realized you could look at changes calculated along branches
on the tree rather than at tip values and that these changes should be independent 
(there are some assumptions about the underlying model of trait evolution you need to make).

A more modern method is phylogenetic generalized least squares (PGLS), where the phylogenetic
relationships between organisms are combined with an assumed model of trait evolution, and
then encoded directly into the covariance matrix of a geralized linear model. These are easy
to implement in R (we will use the `picante` package, but you can also use functions from
`phytools` and `caper`, as well as many other packages).

```{r PGLS}
traits_tree1$Species <- tree1$tip.label
tree1.data <- comparative.data(tree1, traits_tree1, names.col="Species", vcv.dim=2, warn.dropped=TRUE)
model.tree1 <- pgls(Trait1~Trait2, data=tree1.data)
summary(model.tree1)
```

Observe that after correcting for phylogeny, there is no relationship between
our two traits (which is correct). Note that we assume a Brownian Motion model
of trait evolution when fitting our model (which was used to generate the original data so is correct, usually we won't know the underlying model - see below).

If you have discrete traits (e.g., presence/absence or count data) check out the `phylolm` package.

## What Happens If You Use the Wrong Model?

The problem with parametric methods (in general), is that if you assume the wrong
model you can get spurious results. This is also true when picking an evolutionary
model for PGLS. The two most commonly used models are based on common random processes:  
  -  Brownian Motion, which is basically just random non-directional change  
  -  [Ornstein-Uhlenbeck](https://lukejharmon.github.io/pcm/chapter6_beyondbm/), assumes directional change towards some optimum (slighlty more complicated)

For example, below we fit a BM model to trait data generated with an OU process 
and get an incorrect result (there should actually be no association).

```{r OU}
trait_OU_df <- data.frame(Trait1 = fastBM(tree1, mu = 10),
                       Trait2 = fastBM(tree1, mu = 5))
trait_OU_df$Species <- tree1$tip.label
tree1.data <- comparative.data(tree1, trait_OU_df, names.col="Species", vcv.dim=2, warn.dropped=TRUE)
model.tree1 <- pgls(Trait1~Trait2, data=tree1.data)
summary(model.tree1)
```

There's lots of [tutorials](https://lukejharmon.github.io/ilhabela/instruction/2015/07/03/PGLS/) for this type of stuff online. Most packages at least incorporate BM and OU type models.

I'll note that [recently there  has been an emphasis on models that take into account rate shifts in trait evolution across the tree](https://doi.org/10.1093/sysbio/syy031). That is, if the rate of trait evolution
changes in different parts of your phylogeny (e.g., because of a change in mutation rate),
it is possible to get spurious results unless you apply a model that accounts for this.

Finally, you can do a lot of [other things than just look for trait associations](https://www.zoology.ubc.ca/prog/diversitree/)
with this sort of data. For example, you can try and [infer the phenotypes of
now-extinct ancestral species](https://en.wikipedia.org/wiki/Ancestral_reconstruction), or [see if traits influence the rate of diversification of
a lineage](https://academic.oup.com/sysbio/article/56/5/701/1694265).